---
title: "Cleaning Currency Data in R"
author: "Ramon Catalan"
date: "2026-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Data is not always imported into R in a format that is ready for analysis. One common example is salary or other currency data. Values such as "$75,000" may look like numbers, but R can treat them as text because of the dollar sign and comma.

In this code-through, I will show how to identify currency stored as text, clean the values, convert them to numeric data, and use the cleaned values for analysis.

# Create Example Salary Data

First, we will create a small dataset with employee names and salaries.

```{r}
salary_data <- data.frame(
  employee = c("Alex", "Jordan", "Taylor", "Morgan", "Casey"),
  salary = c("$65,000", "$72,500", "$81,250", "$95,000", "$110,500")
)

salary_data
```

# Check the Data Type

Before doing any calculations, it is important to check how R is storing the salary values.

```{r}
class(salary_data$salary)
```

The result shows that the salary column is stored as **character** data instead of numeric data. This happens because the values contain dollar signs and commas.

Because R sees these values as text, we cannot correctly calculate things like the mean or median salary yet.

# What Happens If We Convert It Directly?

A common mistake is trying to convert the salary column directly with `as.numeric()`.

```{r}
as.numeric(salary_data$salary)
```

This does not work correctly because the dollar signs and commas are still part of the text. R cannot interpret values such as "$65,000" as numbers until those extra characters are removed.

This is why cleaning the values first is important.

# Remove Dollar Signs and Commas

Before converting the salary values to numbers, we need to remove the dollar signs and commas.

The `gsub()` function can replace unwanted characters in text.

```{r}
salary_data$salary_clean <- gsub("\\$", "", salary_data$salary)
salary_data$salary_clean <- gsub(",", "", salary_data$salary_clean)

salary_data
```

Now the new `salary_clean` column no longer contains dollar signs or commas. The values look numeric, but they are still stored as character data.


# Convert the Cleaned Values to Numeric

Now that the dollar signs and commas have been removed, we can convert the cleaned salary values to numeric data.

```{r}
salary_data$salary_numeric <- as.numeric(salary_data$salary_clean)

class(salary_data$salary_numeric)

salary_data
```

The new `salary_numeric` column is stored as numeric data. We can now use these values in calculations and statistical summaries.

# Create a Reusable Function

If we need to clean many currency values, it is easier to create a function instead of repeating the same steps each time.

```{r}
convert_currency <- function(x)
{
  x <- gsub("\\$", "", x)
  x <- gsub(",", "", x)
  x <- as.numeric(x)
  return(x)
}
```

Now we can use the function on the original salary column.

```{r}
salary_data$salary_function <- convert_currency(salary_data$salary)

salary_data
```

The function removes the extra currency symbols and converts the values to numeric data in one step.

# Use the Cleaned Salary Data

Now that the salary values are numeric, we can calculate summary statistics.

```{r}
mean(salary_data$salary_numeric)
median(salary_data$salary_numeric)
min(salary_data$salary_numeric)
max(salary_data$salary_numeric)
```

These calculations would not work correctly when the salary values were stored as character data. After cleaning and converting the values, R can use them for analysis.

# Before and After

It can be helpful to compare the original salary values with the cleaned numeric values.

```{r}
salary_data[, c("employee", "salary", "salary_numeric")]
```

The original `salary` column contains dollar signs and commas, so R stores it as character data. The `salary_numeric` column contains only numbers, which makes it ready for calculations, summaries, and graphs.


# Conclusion

Currency values can look like numbers even when R stores them as text. Before using salary data in calculations, it is important to check the data type and remove extra characters such as dollar signs and commas.

In this tutorial, we used `gsub()` to clean the values, `as.numeric()` to convert them, and a custom `convert_currency()` function to make the process reusable.

Cleaning data types correctly is a small step, but it is important because it helps prevent errors and makes the data ready for analysis.
















